Supermartingales and Termination Analysis of Statistical PCF

### 2020-05-30 probabilisticProgrammingMartingales.tex
Examples (random walk) to illustrate the use of (partial) antitone ranking functions

\paragraph{}
In both of these cases, there are only some values of the rankng function that are semantically important. Define a \emph{partial ranking function} on a closed term $M$ to be a partial function 
$f : Rch(M) \rightharpoonup \mathbb R$ 
%$f : Rch(M) \nrightarrow \mathbb R$ 
such that
%\lo{NOTATION. Common notation for partial function $\rightharpoonup$}
\begin{itemize}
    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
    \item $f$ is defined at $M$.
    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually \akr{It would be nicer for this to be only almost certain, to make this more neatly in-between rankability and ast, but that doesn't actually work because of terms reachable with 0 probability. Of course, astness doesn't imply rankability, so any definition of partial ranking function that includes all Y-past terms wouldn't always extend to a total ranking function.} reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
\end{itemize}
A partial ranking function that is total is just a ranking function. Providing a partial ranking function is essentially part way between providing a ranking function and directly proving almost sure termination.

@@

\section{Antitone-convex ranking functions}

\begin{example}
\label{ex:ac-ranking}
\begin{enumerate}
\item 1D unbiased random walk
\[
M_1 = 
\big(\tY \, f \, n \, . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}\big) \, 1
\]

@@

Programs $M_1$ and $M_3$ are AST.
To prove the existence of antitone-convex ranking functions, we appeal \lo{(hopefully)} to a partial ranking function theorem for antitone-convex ranking functions.

\lo{NOTE. Suppose a function $f : E \to \Real$ from a convex subset $E \subseteq \Real^n$ has continuous second-order partial derivatives. Then it is convex iff its Hessian is positive semidefinite.}

For $M_1$, consider the function $g_1(n) := \frac{2^n - 1}{2^{n-1}}$.
Note that $(g_1 (n))_{n \geq 0}$ satisfies the recurrence relation
\[
z_0 = 0, \qquad
z_n > 1/2 \cdot z_{n-1} - 1/2 \cdot z_{n+1} \quad (n \geq 1)
\]
and set the antitone and convex function $\epsilon_1$ to be 
\[
x \mapsto \big(g_1(x) - 1/2 \cdot g_1(x-1) - 1/2 \cdot g_1(x+1)\big) = \frac{1}{2^{x+1}}.
\]

For $M_3$, we use the function $g_3(n) := \frac{3^n - 1}{3^{n-1}}$, and set $\epsilon_3$ to be 
\[
x \mapsto \big(g_3(x) - 2/3 \cdot g_3(x-1) - 1/3 \cdot g_3(x+2) \big) = 
\frac{28}{3^{x+2}}.
\]

>>>>

\paragraph{}
In both of these cases, there are only some values of the ranking function that are semantically important. Define a \emph{partial ranking function} on a closed term $M$ to be a partial function 
$f : Rch(M) \rightharpoonup \mathbb R$ 
%$f : Rch(M) \nrightarrow \mathbb R$ 
such that
%\lo{NOTATION. Common notation for partial function $\rightharpoonup$}
\begin{itemize}
    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
    \item $f$ is defined at $M$.
    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually \akr{It would be nicer for this to be only almost certain, to make this more neatly in-between rankability and ast, but that doesn't actually work because of terms reachable with 0 probability. Of course, astness doesn't imply rankability, so any definition of partial ranking function that includes all Y-past terms wouldn't always extend to a total ranking function.} reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
\end{itemize}
A partial ranking function that is total is just a ranking function. Providing a partial ranking function is essentially part way between providing a ranking function and directly proving almost sure termination.

@@

\section{Antitone ranking functions}

\begin{example}
\label{ex:ac-ranking}
\begin{enumerate}
\item 1D unbiased random walk
\[
M_1 = 
\big(\tY \, f \, n \, . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}\big) \, 10
\]

@@

Programs $M_1$ and $M_3$ are AST.
To construct antitone ranking functions, we appeal \lo{(hopefully)} to a partial ranking function theorem for antitone ranking functions.
\iffalse
\lo{NOTE. Suppose a function $f : E \to \Real$ from a convex subset $E \subseteq \Real^n$ has continuous second-order partial derivatives. Then it is convex iff its Hessian is positive semidefinite.}
\fi
\iffalse
For $M_1$, consider the function $g_1(n) := \frac{2^n - 1}{2^{n-1}}$.
\fi

For $M_1$, define the function $g_1 : \nnReal \to [0, 2)$ by $g_1(x) = 2 - \frac{1}{2^{x-1}}$.
Note that $(g_1 (n))_{n \in \omega}$ satisfies the recurrence relation
\[
\left\{
\begin{array}{rll}
z_0 &=& 0\\
n \geq 1, \quad z_n &>& \dfrac{1}{2} \, \big( z_{n-1} + z_{n+1}\big).
\end{array}
\right.
\]
Set the antitone function $\epsilon_1 (y) :=  \frac{2-y}{4}$.
Then
\begin{equation}
\epsilon_1(g_1(x)) = g_1(x) - \dfrac{1}{2} \, \big(g_1(x-1) + g_1(x+1)\big) = \dfrac{1}{2^{x+1}}
\label{eqn:epsilon g M1}
\end{equation}
(Since $g_1$ is bijective, with inverse $g_1^{-1}(y) = 1 - \log (2-y)$, we can obtain $\epsilon_1(y)$ from \Cref{eqn:epsilon g M1}.) 
%We have $1/2 \cdot \big(g_1(x-1) + g_1(x+1)\big) = 2 - \frac{5}{2^{x+1}}$.

Using shorthand 
$\Theta_1 = \tY \, f \, n \, . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
we can define a partial antitone ranking function $f_1 : \mathit{Rch}(M_1) \to \nnReal$ with antitone function $\epsilon_1$ defined as follows:
\begin{align*}
{\Theta_1} \, n 
&\mapsto 
g_1(n)
\\
\tif{n = 0}{0}{\Theta_1 \, (n - 1) \oplus_{1/2} \Theta_1 \, (n + 1)}
&\mapsto
\dfrac{1}{2} \, \big(g_1(n-1) + g_1(n+1)\big)
\\
0 &\mapsto 0
\end{align*}

For $M_3$, we use the function $g_3(x) := 3 - \frac{1}{3^{x-1}}$ and the antitone function $\epsilon_3(y) := \frac{84-28y}{27}$. 
Then
\[
\epsilon_3(g_3(x)) = g_3(x) - \dfrac{1}{3}\big(2 \cdot g_3(x-1) + \cdot g_3(x+2) \big) = 
\frac{28}{3^{x+2}}.
\]
Following $M_1$ above, we can define a partial antitone ranking function $f_3 : \mathit{Rch}(M_3) \to \nnReal$ with $\epsilon_3$ as the antitone function.

\begin{example}[Continuous random walk]\label{ex:raven complex}
In SPCF we can construct a function whose argument changes by a random amount at each recursive call:
\[
\underbrace{\big
(\tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample))} \big)}_{\Theta} \, 10
\]
We can construct a partial antitone ranking function $f$ as follows:
\begin{align*}
\Theta \, l 
&\mapsto 
l
\\
\tif{l \leq 0}{0}{\Theta \, (l - \tsample)}
&\mapsto
l - 1/2
\\
0 &\mapsto 0
\end{align*}
with the constant antitone function $\epsilon(y) = 1/2$.

For a more complex example, consider the following ``continuous random walk''
\[
\underbrace{\big
(\tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample) \oplus_{1/2} f(x - \tsample)} \big)}_{\Xi} \, 10
\]
Let $g(x) := \sqrt{2} - \frac{1}{2^{x-\frac{1}{2}}}$ and let $\epsilon$ be the function specified by
\[
\epsilon(g(x)) = g(x) - \dfrac{1}{2}\Big(g(x - \dfrac{1}{2}) + \cdot g(x + \dfrac{1}{2}) \Big) = 
\dfrac{3 - 2\sqrt{2}}{2^{x+1}}.
\]
We define a partial antitone ranking function as follows:
\begin{align*}
\Xi \, l 
&\mapsto 
g(l)
\\
\tif{l \leq 0}{0}{\Xi \, (l - \tsample) \oplus_{1/2} \Xi \, (l + \tsample)}
&\mapsto
\frac{1}{2}\big(g(l - \frac{1}{2}) + \cdot g(l + \frac{1}{2}) \big)
\\
r \leq 0, \quad r &\mapsto 0
\end{align*}
\end{example}
